---
title: "PA1_template.Rmd"
author: "jspp"
date: "10/12/2014"
output: html_document
---
# 
# 
#1-Loading and Tranforming Data
###1.1- Loading data, including download
data is downloaded in zip format, then extracted and then read into R

```{r }
download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",
              "represearch/factivitydata.zip")
unzip("represearch/factivitydata.zip", exdir = "represearch/.")
activitydata <- (read.csv("represearch/activity.csv"))
```

###1.2- Transforming dates into date format
initially dates are read as factor, here the class is changed
```{r }
activitydata$date <- as.Date (activitydata$date)
```

#2 Total Number of Steps every Day
generates a variable that states the total steps, mean steps(by 5 min interval)
and the number of missing values every day

```{r }
daily.steps.calc <- function()  {
    total.steps <- tapply(activitydata$steps, activitydata$date, sum, na.rm = TRUE)
    step.means <- tapply(activitydata$steps, activitydata$date, mean, na.rm = TRUE)
    NA.count <- sapply(tapply(activitydata$steps, activitydata$date, is.na, 
                              simplify = FALSE),sum, simplify = TRUE)
    daily.steps <<- as.data.frame(cbind(total.steps, step.means, NA.count))
}
daily.steps.calc()
```
##2.1 Histogram
loads ggplot2 and plots the histogram
```{r}
library(ggplot2)
ggplot(daily.steps, aes (x=total.steps)) +
        geom_histogram(binwidth = (max(daily.steps$total.steps)/8), 
        colour = "black", fill= "white")
```

###2.2 Mean and median 
reports Mean, median and a couple of extra things for the total steps every day

```{r}
summary(daily.steps$total.steps)
```

#3 Average number of Steps per 5 min Interval

###Calculating the averages

```{r }
steps.interval <- as.data.frame(tapply(activitydata$steps, activitydata$interval,
                         mean, na.rm = TRUE))
steps.interval <- cbind(as.numeric(rownames(steps.interval)),steps.interval)
colnames(steps.interval) <- c("Interval","Mean.Steps.in.Interval") 
```

###plot
once aggain ggplot2 is loaded and the plot is generated
```{r}
library(ggplot2)
g <- ggplot(steps.interval , aes(Interval,Mean.Steps.in.Interval)) 
g + geom_line()
```

###maximum
reports the maximum
```{r}
steps.interval[which.max(steps.interval$Mean.Steps.in.Interval),]
```

#4 Missing Values

###4.1 SumNA's
sums all the NA values we calculated before, in section 2

```{r}
sum(daily.steps$NA.count)
```

Since data missing corresponds to complete days
one cannot fill thw data with estimates from the day,
therefor the approach whould be 
using the mean for the 5 min Interval form other values
    
###4.2 Filling data in new dataset

a copy of the data is generated, then means by each interval are calculated, 
finaly taking means for the respective interval data is filled in the NA spaces

```{r }
activitydatafill <- activitydata
activitydatafill$steps[is.na(activitydata$steps)] <- 
        (tapply(activitydata$steps, activitydata$interval,mean, na.rm = TRUE))[
        as.factor(activitydata$interval[is.na(activitydata$steps)])]
```

###4.3 Total Number of Steps every Day
Equivalent to section 2 with the filled data

```{r }
daily.steps.calc.fill <- function()  {
    total.steps <-  tapply(activitydatafill$steps, 
                            activitydatafill$date, sum, na.rm = TRUE)
    step.means <-   tapply(activitydatafill$steps, 
                            activitydatafill$date, mean, na.rm = TRUE)
    NA.count <- sapply(tapply(activitydatafill$steps, activitydatafill$date, 
                              is.na, simplify = FALSE),sum, simplify = TRUE)
    daily.steps.fill <<- as.data.frame(cbind(total.steps, step.means, NA.count))
}
daily.steps.calc.fill()
```

###4.4 -Makes Histogram

```{r}
library(ggplot2)
ggplot(daily.steps.fill, aes (x=total.steps)) +
    geom_histogram(binwidth = (max(daily.steps.fill$total.steps)/8), 
                   colour = "black", fill= "white")
```

###4.5 reports mean, median and a couple of extra things

```{r}
summary(daily.steps.fill$total.steps)
```

compare with the data obtained without filling

```{r}
summary(daily.steps$total.steps)
```

it is possible to see there are differences in the two data sets looking at the
mean, minimum value and 1st quartile.

#5 Weekday-Weekend difference
###5.1 Addition of extra factor column
a collumn "weekday" is added with the factors "weekday" and "weekend" to state
if the corresponding interval occurs on a weekday or a weekend
```{r }
weekdayaddition <- function() {
    tweekdays <- c("Monday","Tuesday","Wednesday","Thursday","Friday")
    tweekends <- c("Saturday","Sunday")
    weekday <- weekdays(activitydatafill$date)
    weekday[weekday %in% tweekdays] <- "weekday"
    weekday[weekday %in% tweekends] <- "weekend"
    activitydatafill <<- cbind(activitydatafill, weekday)
}
weekdayaddition()
```

### 5.2 Plot Grid between weekdays and weekends interval mean

First data is separated insto weekdays and weekends, then means are calculated
and finally data is merged aggain to be exported to the global environment

Then ggplo2 is loaded and the graph is generated

```{r cache = TRUE}
interval.mean.by.weekday <- function() {
    weekdays <- activitydatafill[activitydatafill$weekday == "weekday",]
    weekends <- activitydatafill[activitydatafill$weekday == "weekend",]
    step.inter.fill.wd <- tapply(weekdays$steps,weekdays$interval, mean)
    step.inter.fill.wd <- data.frame(step.inter.fill.wd, 
                                     as.numeric(rownames(step.inter.fill.wd)),
                                     "weekdays")
    step.inter.fill.ed <- tapply(weekends$steps,weekends$interval, mean)
    step.inter.fill.ed <- data.frame(step.inter.fill.ed, 
                                     as.numeric(rownames(step.inter.fill.ed)),
                                     "weedkend")
    colnames(step.inter.fill.wd)<-c("steps.mean", "interval","weekday")
    colnames(step.inter.fill.ed)<-c("steps.mean", "interval","weekday")
    step.inter.fill.wd$step<-as.numeric(step.inter.fill.wd$step)
    step.inter.fill.ed$step<-as.numeric(step.inter.fill.ed$step)
    intervalmeans <<- rbind(step.inter.fill.wd,step.inter.fill.ed)
}
interval.mean.by.weekday()

library(ggplot2)
pg <- ggplot(intervalmeans, aes(x = interval, y = steps.mean))+
          geom_line() + facet_grid(weekday ~ .)
pg
```

